Notes on $\etarho$

### 2009-02-05 docs/eta_notes/eta.tex
switch over to book format for the master document -- this makes the sectioning more natural
add MarginPars
fixed the etarho flow chart to reflect a time-centered eta
make the definitions of \half and \myhalf consistent throughout the notes
addressed some marginpars, put in section labels with proper referencing in the flowchart
put some notes about eta in spherical coordinate in here -- they were originally in paper IV

\documentclass[11pt]{article}

\tolerance=600

\usepackage{amsmath,color}

% Margins
\usepackage[lmargin=0.5in,rmargin=1.5in,tmargin=0.5in,bmargin=0.5in]{geometry}

\def\half   {\frac{1}{2}}

\def\eb         {{\bf e}}
\def\ib         {{\bf i}}
\def\Ub         {{\bf U}}

\newcommand{\Ubt}{\widetilde{\Ub}}

\def\etarho     {\eta_\rho}
\def\etarhoec   {\etarho^{\rm ec}}
\def\etarhocc   {\etarho^{\rm cc}}
\def\etarhoflux {\etarho^{\rm flux}}
\def\divetarho  {\nabla\cdot(\etarho\eb_r)}

\title{Notes on $\etarho$}

\setlength{\marginparwidth}{1.0in}
\newcommand{\MarginPar}[1]{\marginpar{%
\vskip-\baselineskip %raise the marginpar a bit
\raggedright\tiny\sffamily
\hrule\smallskip{\color{red}#1}\par\smallskip\hrule}}

\begin{document}

\maketitle
\tableofcontents
\clearpage

@@

\section{Flow Chart}
\begin{enumerate}
\item Enter {\tt advance} with $[\etarhoec, \etarhocc, \divetarho]^{n-\half}$.
\item Call {\tt make\_w0}.  The spherical version uses uses $\etarhoec$ and $\etarhocc$.
\item Call {\tt density\_advance}.  The plane-parallel version computes $\etarhoflux$.
\item Call {\tt correct\_base}, which uses $\divetarho$.
\item Compute $[\etarhoec, \etarhocc, \divetarho]^{n+\half,*}$.
\item Call {\tt make\_psi}.  The plane-parallel version uses $\etarhocc$.
\item Call {\tt make\_w0}.  The spherical version uses uses $\etarhoec$ and $\etarhocc$.
\item Call {\tt density\_advance}.  The plane-parallel version computes $\etarhoflux$.
\item Call {\tt correct\_base}, which uses $\divetarho$.
\item Compute $[\etarhoec, \etarhocc, \divetarho]^{n+\half}$.
\item Call {\tt make\_psi}.  The plane-parallel version uses $\etarhocc$.
\end{enumerate}

\section{Computing $\etarhoec, \etarhocc$, and $\divetarho$}
This is done in {\tt make\_eta.f90}.

\subsection{Plane-Parallel}
We first compute a radial edge-centered multifab, $\eta_{\rho}^{\rm flux}$, using
\begin{equation}
\eta_{\rho,\ib+\half\eb_r}^{\rm flux} = \left[\left(\Ubt_{\ib+\half\eb_r}^{n+\half}\cdot\eb_r\right) + w_{0,r+\half}^{n+\half}\right] \rho_{\ib+\half\eb_r}^{n+\half} - w_{0,r+\half}^{n+\half}\rho_{0,r+\half}^{n+\half, {\rm pred}}
\end{equation}
$\etarhoec$ is the edge-centered ``average'' value of $\eta_{\rho}^{\rm flux}$,
\begin{equation}
\eta_{\rho,r+\half}^{\rm ec} = \overline{\eta_{\rho,\ib+\half\eb_r}^{\rm flux}}
\end{equation}
$\etarhocc$ is a cell-centered average of $\etarhoec$,
\begin{equation}
\eta_{\rho,r}^{\rm cc} = \frac{\eta_{\rho,r+\half}^{\rm ec} + \eta_{\rho,r-\half}^{\rm ec}}{2}.
\end{equation}
$\divetarho$ is the cell-centered divergence of $\etarhoec$,
\begin{equation}
\left[\divetarho\right]_r = \frac{\eta_{\rho,r+\half}^{\rm ec} - \eta_{\rho,r-\half}^{\rm ec}}{\Delta r}.
\end{equation}

\subsection{Spherical}
First, construct $\eta_{\rho}^{\rm cart} = [\rho'(\Ubt\cdot\eb_r)]^{n+\half}$ using:
\begin{equation}
\left[\frac{\rho^n+\rho^{n+1}}{2}-\left(\frac{\rho_0^n+\rho_0^{n+1}}{2}\right)^{\rm cart}\right] \sum_d\left(\frac{\Ubt_{\ib+\half\eb_d}^{n+\half}\cdot\eb_d+\Ubt_{\ib-\half\eb_d}^{n+\half}\cdot\eb_d}{2}\right)n_d.
\end{equation}
Then, $\etarhocc$ is the cell-centered average of $\eta_{\rho}^{\rm cart}$,
\begin{equation}
\etarhocc = \overline{\eta_{\rho}^{\rm cart}}.
\end{equation}
On interior faces, $\etarhoec$ is the average of $\etarhocc$,
\begin{equation}
\eta_{\rho,r-\half}^{\rm ec} = \frac{\eta_{\rho,r-1}^{\rm cc} + \eta_{\rho,r}^{\rm cc}}{2}.
\end{equation}
At the upper and lower boundaries, we use
\begin{eqnarray}
\eta_{\rho,-\half}^{\rm ec} &=& 0, \\
\eta_{\rho,{\rm nr}-\half}^{\rm ec} &=& \eta_{\rho,{\rm nr}-1}^{\rm cc}.
\end{eqnarray}
$\divetarho$ is constructed from the cell-centered average of $\rho'^{n+1} = \rho^{n+1} - \rho_0^{n+1}$ as
\begin{equation}
\divetarho = -\overline{\rho'^{n+1}}/\Delta t
\end{equation}
\section{Using $\etarhoec$}
\subsection{Plane-Parallel}
NOT USED.
\subsection{Spherical}
In {\tt make\_w0}, $\etarhoec$ is used in the construction of the RHS for the $\delta w_0$ equation.
\section{Using $\etarhocc$}
\subsection{Plane-Parallel}
In {\tt make\_psi}, $\psi = \etarhocc g$.
\subsection{Spherical}
In {\tt make\_w0}, $\etarhocc$ is used in the construction of the RHS for the $\delta w_0$ equation.
\section{Using $\divetarho$}
\subsection{Plane-Parallel}
In {\tt correct\_base}, $\rho_0^{n+1} = \rho_0^{n+1} - \Delta t \divetarho$.
\subsection{Spherical}
In {\tt correct\_base}, $\rho_0^{n+1} = \rho_0^{n+1} - \Delta t \divetarho$. \\
\end{document}

>>>>

@@

\section{The Mixing Term, $\etarho$}

The base state evolves in response to heating and mixing in the star.
The density evolution is governed by
\begin{eqnarray}
\frac{\partial \rho_0}{\partial t} &=& - 
 \nablab \cdotb \left( \rho_0 w_0 \er \right)
- \nablab \cdotb \left( \etarho \er \right) \enskip ,
\label{eq:rho0upd_new}
\end{eqnarray}
with
\begin{equation}
\etarho(r) = \overline{\left(\rhop \ut \cdot \er \right)} = \frac{1}{A(\Omega_H)} 
 \int_{\Omega_H}  (\rhop \ut \cdot \er ) \; dA \enskip , \label{eq:eta}
\end{equation}
designed to keep the average value of the full density, $\rho$, over a
layer of constant radius in the star equal to $\rho_0$.  To complete
the update of the base state, we need evolution equations for the
pressure, $p_0$, and velocity, $w_0$.  For spherical geometry, the
derivation of $w_0$ constraint equation is shown in
Appendix~\ref{app:gravity}, resulting in the following system
\begin{eqnarray}
w_0 &=& \ow + \delta w_0 \\
\frac{1}{r^2} \frac{\partial}{\partial r} \left (r^2 \ow \right ) &=& \Sbar \\
\frac{\partial}{\partial r} \left[ \frac{\gammabar p_0}{r^2} \frac{\partial}{\partial r} (r^2 \dw) \right] &=& - \frac{g}{r^2} \frac{\partial (r^2 \etarho)}{\partial r} - \frac{4 (\ow + \dw) \rho_0 g}{r} 
- 4 \pi G \rho_0 \etarho \label{eq:dw0constraint}
\end{eqnarray}

In paper~III, we introduced a mixing term, $\etarho$, to the density
evolution equation (eq.~[\ref{eq:rho0upd_new}]), with the objective of
keeping the base state density equal to the average of the density
over a layer.  For a spherical base state, it is best to define the 
average in terms of spherical coordinates,
\begin{equation}
\overline{q} = \frac{1}{4\pi} \int_{\Omega_H} q(r,\theta,\phi) \; d\Omega
\end{equation}
where $\int_{\Omega_H} d\Omega = 4\pi$ represents the integral over
the spherical $\theta$ and $\phi$ angles at constant radius.

Recall that if initially $ \overline{\rhop} = 1 /
{4\pi} \int_{\Omega_H} \rhop \; d\Omega = 0,$ 
\MarginPar{this section is here temporarily, just to better understand
what's going on here.  These notes came from an early version of
paper~III} 
there is no guarantee that $\overline{\rhop} = 0$ will hold at later
time.  To see this we start with the equation for $\rho'$, 
\begin{equation}
\frac{\partial \rhop}{\partial t} = - \nablab \cdotb (\rho_0 \ut)
-\nablab \cdotb (\rhop w_0 \er) - \nablab \cdotb (\rhop \ut )
\label{eq:rhoprimeold} \enskip ,
\end{equation}
rewritten by moving one term to the left-hand-side:
\begin{equation}
\frac{\partial \rhop}{\partial t} + \nablab \cdotb (\rhop \ut) 
= - \nablab \cdotb (\rho_0 \ut)
  -\nablab \cdotb (\rhop w_0 \er) \enskip .
\end{equation}
We integrate this over a spherical shell of thickness $2h$ at radius $r_0$, 
$\Omega_H \times (r_0-h, r_0+h)$, and normalize by the integration
volume, which is $\sim 4\pi r_0^2 \, 2h$ for small $h$, to obtain:
\begin{eqnarray}
\frac{1}{4\pi r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} r^2 \; dr \int_{\Omega_H} 
    \left(\frac{\partial \rhop}{\partial t} + \nabla \cdotb ( \rhop \ut ) \right) \; d\Omega 
&=& - \frac{1}{4\pi r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} r^2 \; dr 
\int_{\Omega_H} \nabla \cdotb \left( \rho_0 \ut + \rhop w_0 \er \right) \; d\Omega \nonumber  \\
&=& \left .  - \frac{1}{4\pi r_0^2 \, 2h}  \int_{\Omega_H} 
\left [ \rho_0  (\ut \cdot \er)  + \rhop w_0 \right] r^2
  \; d\Omega  \; \right  |_{r_0-h}^{r_0+h} \nonumber  \\
&=& 0 \nonumber \enskip ,
\end{eqnarray} 
where we have used the divergence theorem in spherical coordinates to transform
the volume integrals into area integrals over $\Omega_H.$
We see that both of the terms on the right hand side disappear. The first,
$\int_{\Omega_H} \rho_0 (\ut \cdot \er) d\Omega = 0$ follows from the definition of $\ut,$
and the second, $\int_{\Omega_H} \rhop w_0 d\Omega = w_0 \int_{\Omega_H} \rhop d\Omega = 0$
follows from our initial assumption.  
%\begin{eqnarray}
%\frac{1}{4\pi r_0^2 \, 2h} \frac{\partial}{\partial t} 
%  \int_{r_0-h}^{r_0+h} \int_{\Omega_H}  \rhop \; r^2 dr \; d\Omega &=&  
%\frac{1}{4\pi r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} \int_{\Omega_H} \nabla \cdotb \left( \rho_0 \ut + \rhop w_0 \er \right)
%\; r^2 dr \; d\Omega &=&
%\\
%&=& \left .  - \frac{1}{4\pi r_0^2 \, 2h} \int_{\Omega_H} 
% \rhop (\ut \cdot \er)  \; r^2 d\Omega  \right  |_{r_0-h}^{r_0+h}
%\enskip ,
%\end{eqnarray} 
Now expanding the remaining terms and taking the limit as $h \rightarrow 0,$ we can write 
\begin{eqnarray} 
0 &=& \lim_{h\rightarrow 0} \frac{1}{4\pi r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} r^2 \; dr \int_{\Omega_H} 
    \left(\frac{\partial \rhop}{\partial t} + \nabla \cdotb ( \rhop \ut ) \right) \; d\Omega \nonumber \\
&=& \frac{\partial}{\partial t} \left( \lim_{h\rightarrow 0} \frac{1}{4\pi r_0^2} \frac{1}{2h} 
  \int_{r_0-h}^{r_0+h} r^2 \; dr \int_{\Omega_H}  \rhop \;d\Omega \right) +
\lim_{h\rightarrow 0} \left( \frac{1}{4\pi r_0^2 \, 2h}  
  \int_{r_0-h}^{r_0+h} r^2 \; dr \int_{\Omega_H}  \nabla \cdotb ( \rhop \ut ) \; d\Omega \right) \nonumber \\
&=& \frac{\partial}{\partial t} \left(\frac{1}{4\pi} \int_{\Omega_H}  \rhop \; d\Omega \right) +
\lim_{h\rightarrow 0} \left( \left .  \frac{1}{4\pi r_0^2 \, 2h}  
\left [ \int_{\Omega_H} \rhop (\ut \cdot \er) \; d\Omega  \right]  r^2\; \right |_{r_0-h}^{r_0+h} \right) \nonumber  \\
&=&  \frac{\partial}{\partial t} \overline{\rhop} + \lim_{h\rightarrow 0} \left( \left .  \frac{1}{r_0^2 \, 2h}
\overline{[\rhop (\ut \cdot \er)]} \; r^2 \; \right |_{r_0-h}^{r_0+h} \right) \nonumber  \\
&=&  \frac{\partial}{\partial t} \overline{\rhop} + 
\lim_{h\rightarrow 0} \frac{1}{r_0^2 \, 2h} \int_{r_0-h}^{r_0+h} \nabla \cdot \left[\overline{\left(\rhop\ut \cdot \er\right)} \er \right]\;  r^2 \; dr \nonumber \\
&=&  \frac{\partial}{\partial t} \overline{\rhop} + \nabla \cdot \left[ \overline{\left(\rhop \ut \cdot \er\right)} \er \right]
\end{eqnarray} 
again using the divergence theorem, extracting the time derivative from the spatial integral,
and switching the order of operations as appropriate.

In short, 
\begin{equation}
\frac{\partial}{\partial t} \overline{\rhop} = 
- \nabla \cdot \left [ \overline{\left(\rhop \ut \cdot \er\right)} \, \er \right ] = 
- \nabla \cdot ( \etarho \er )
 \enskip .\label{eq:rhopbar} 
\end{equation}

We need both $\etarho$ alone and its divergence for the
various terms in the construction of $w_0$ and the correction to
$\rho_0$.  The quantity $\etarho$ is edge-centered on our grid, and
for Cartesian geometries, we constructed it by averaging the
appropriate fluxes through the grid boundaries.  For a spherical base
state, this does not work, since the spherical shells do not line up
with the Cartesian grid boundaries.

Therefore, we take a different approach.  We compute $\etarho$ by
constructing the quantity $\rhop \ut \cdot \er$ in each cell, and then
use our average routine to construct a 1-d, cell-centered
$\eta_{\rho,r}$ (this is essentially numerically solving the integral
in Eq.~[\ref{eq:eta}]).  The edge-centered values of $\etarho$,
$\eta_{\rho,r+1/2}$ are then constructed by simple
averaging: \MarginPar{should we be doing a volume-weighted average?}
\begin{equation}
\eta_{\rho,r+1/2} = \frac{\eta_{\rho,r} + \eta_{\rho,r+1}}{2} \enskip .
\end{equation}

Instead of differencing $\eta_{\rho,r+1/2}$ to construct the 
divergence, we instead use equation~(\ref{eq:rhopbar}) directly, by writing:
\begin{equation}
\left [ \nabla \cdot (\etarho \er ) \right ]^{n+1/2}
= - \frac{\overline{\rhop^{n+1}} - \overline{\rhop^n}}{\Delta t}
= - \frac{\overline{\rhop^{n+1}}}{\Delta t} \enskip ,
\end{equation}
where we have made use of the fact that $\overline{\rhop^n} = 0$ by construction.

%-----------------------------------------------------------------------------
% flow chart
%-----------------------------------------------------------------------------
\section{$\eta$ Flow Chart}
\begin{enumerate}
\item Enter {\tt advance\_timestep} with $[\etarhoec, \etarhocc, \divetarho]^{n-\myhalf}$.
\item Call {\tt make\_w0}.  The spherical version uses uses $\etarhoec$ and $\etarhocc$.
\item Call {\tt density\_advance}.  The plane-parallel version computes $\etarhoflux$.
\item Call {\tt make\_etarho} to compute $[\etarhoec, \etarhocc, \divetarho]^{n+\myhalf,*}$
\item Call {\tt make\_psi}.  The plane-parallel version uses $\etarhocc$.
\item Call {\tt correct\_base}, which uses $\divetarho$.
\item Call {\tt make\_w0}.  The spherical version uses uses $\etarhoec$ and $\etarhocc$.
\item Call {\tt density\_advance}.  The plane-parallel version computes $\etarhoflux$.
\item Call {\tt make\_etarho} to compute $[\etarhoec, \etarhocc, \divetarho]^{n+\myhalf}$
\item Call {\tt make\_psi}.  The plane-parallel version uses $\etarhocc$.
\item Call {\tt correct\_base}, which uses $\divetarho$.
\end{enumerate}

%-----------------------------------------------------------------------------
% computing
%-----------------------------------------------------------------------------
\section{Computing $\etarhoec, \etarhocc$, and $\divetarho$}
This is done in {\tt make\_eta.f90}.

\subsection{Plane-Parallel}
We first compute a radial edge-centered multifab, $\eta_{\rho}^{\rm flux}$, using
\begin{equation}
\eta_{\rho,\ib+\myhalf\eb_r}^{\rm flux} = \left[\left(\Ubt_{\ib+\myhalf\eb_r}^{n+\myhalf}\cdot\eb_r\right) + w_{0,r+\myhalf}^{n+\myhalf}\right] \rho_{\ib+\myhalf\eb_r}^{n+\myhalf} - w_{0,r+\myhalf}^{n+\myhalf}\rho_{0,r+\myhalf}^{n+\myhalf, {\rm pred}}
\end{equation}
$\etarhoec$ is the edge-centered ``average'' value of $\eta_{\rho}^{\rm flux}$,
\begin{equation}
\eta_{\rho,r+\myhalf}^{\rm ec} = \overline{\eta_{\rho,\ib+\myhalf\eb_r}^{\rm flux}}
\end{equation}
$\etarhocc$ is a cell-centered average of $\etarhoec$,
\begin{equation}
\eta_{\rho,r}^{\rm cc} = \frac{\eta_{\rho,r+\myhalf}^{\rm ec} + \eta_{\rho,r-\myhalf}^{\rm ec}}{2}.
\end{equation}
$\divetarho$ is the cell-centered divergence of $\etarhoec$,
\begin{equation}
\left[\divetarho\right]_r = \frac{\eta_{\rho,r+\myhalf}^{\rm ec} - \eta_{\rho,r-\myhalf}^{\rm ec}}{\Delta r}.
\end{equation}

\subsection{Spherical}\label{Sec:eta Spherical}
First, construct $\eta_{\rho}^{\rm cart} =
[\rho'(\Ubt\cdot\eb_r)]^{n+\myhalf}$ using:
\begin{equation}
\left[\frac{\rho^n+\rho^{n+1}}{2}-\left(\frac{\rho_0^n+\rho_0^{n+1}}{2}\right)^{\rm cart}\right] \sum_d\left(\frac{\Ubt_{\ib+\myhalf\eb_d}^{n+\myhalf}\cdot\eb_d+\Ubt_{\ib-\myhalf\eb_d}^{n+\myhalf}\cdot\eb_d}{2}\right)n_d.
\end{equation}
Then, $\etarhocc$ is the cell-centered average of $\eta_{\rho}^{\rm cart}$,
\begin{equation}
\etarhocc = \overline{\eta_{\rho}^{\rm cart}}.
\end{equation}
On interior faces, $\etarhoec$ is the average of $\etarhocc$,
\begin{equation}
\eta_{\rho,r-\myhalf}^{\rm ec} = \frac{\eta_{\rho,r-1}^{\rm cc} + \eta_{\rho,r}^{\rm cc}}{2}.
\end{equation}
At the upper and lower boundaries, we use
\begin{eqnarray}
\eta_{\rho,-\myhalf}^{\rm ec} &=& 0, \\
\eta_{\rho,{\rm nr}-\myhalf}^{\rm ec} &=& \eta_{\rho,{\rm nr}-1}^{\rm cc}.
\end{eqnarray}
$\divetarho$ is constructed from the cell-centered average of
$\rho'^{n+1} = \rho^{n+1} - \rho_0^{n+1}$ as
\begin{equation}
\divetarho = -\overline{\rho'^{n+1}}/\Delta t
\end{equation}

%-----------------------------------------------------------------------------
% etarho_ec
%-----------------------------------------------------------------------------
\section{Using $\etarhoec$}
\subsection{Plane-Parallel}
NOT USED.

\subsection{Spherical}
In {\tt make\_w0}, $\etarhoec$ is used in the construction of the RHS
for the $\delta w_0$ equation.

%-----------------------------------------------------------------------------
% etarho_cc
%-----------------------------------------------------------------------------
\section{Using $\etarhocc$}
\subsection{Plane-Parallel}
In {\tt make\_psi}, $\psi = \etarhocc g$.

\subsection{Spherical}
In {\tt make\_w0}, $\etarhocc$ is used in the construction of the RHS
for the $\delta w_0$ equation.

%-----------------------------------------------------------------------------
% div etarho
%-----------------------------------------------------------------------------
\section{Using $\divetarho$}
\subsection{Plane-Parallel}
In {\tt correct\_base}, $\rho_0^{n+1} = \rho_0^{n+1} - \Delta t \divetarho$.

\subsection{Spherical}
In {\tt correct\_base}, $\rho_0^{n+1} = \rho_0^{n+1} - \Delta t \divetarho$. \\
